# Using and Building Functions {#sec-using-and-building-functions}

```{r}
#| label: load-prereqs
#| echo: false
#| message: false

# knitr options
source("_common.R")

```

## Overview 

This section will discuss R functions. Functions are the engines that define how to DO things with your data objects in R. If data objects are the "nouns", then functions are the "verbs" of the R language. 

Why build functions?

1. Because it is fun (for us nerdy people).

2. Because we find ourselves doing the same task multiple times throughout our code, and realize that each iteration is mostly the same process using slightly different values or variables. 

3. Because we want to avoid copy/paste related errors.

4. Because we want to make our code more readable.

This will discuss the basic structure and components of a function, how a function is created, passing parameters implicitly versus explicitly in a function call, name masking in nested environments, and infix operators.

Below are links to supplementary resources about building functions in R: 

- [Advanced R Functions Chapter](http://adv-r.had.co.nz/Functions.html)
- [Additional Functions Reading from R for Data Science E-book](https://r4ds.hadley.nz/functions)

## R Function Basics - Components 

>``To understand computations in R, two slogans are helpful: Everything that exists is an object. Everything that happens is a function call.''
  — John Chambers

Formals: The defined parameters of a function. Think of these as the set of data objects and options that can be defined by the programmer when a function is used.

For these exercises we will consider the following functions:

- `mean()` from base R
- `ggplot()` from the ggplot2 package
- `%>%` piping symbol from the dplyr package

```{r}
#| label: function-calls 

library(tidyverse)

# a function without parentheses prints what it is
# -> with the parenthesis, it tells R to run it
mean
ggplot

```

Pipe

- This is actually quite a complex function (yes, a function) (things like this are called infix operator). Infix operators go in between arguments.

- Can surround an infix operator with tick marks to see what this function is actually doing.

```{r}
#| label: pipe

# infix operator: pipe
`%>%`

```

- Another example is `+`. This is a primitive function, which just means that R doesn't evaluate it. It sends it down and is evaluated in C (because it is more efficient).

```{r}
#| label: addition

# infix operator as function
`+`(2, 3)

```

### Formals

What are the formals for these functions?

- `?formals()`: Formals are things that we define to change the behavior of the outcome.

```{r}
#| label: formals

# view formals
formals(`%>%`)
formals(mean)

```

- `...` are a special (very flexible, catch all) type of formal argument that we can put non-necessary arguments.

- For example, with mean `mean(x, na.rm = TRUE)`, this saves the value `TRUE` to na.rm  argument and saves that value as the function is run.

### Body 

The lines of code that will be run when the function is called. The body uses the formals and usually returns a result of some kind.

```{r}
#| label: body

# body()
body(`+`)

# -> returns null because the function isn't run in R

body(ggplot)

```

- Note that the code in the body is never commented because R strips away the comments when running to make its more compact. When we write functions, we add comments so the humans know what it does, but when it is stored they are removed to take up less space.

### Environment 

The environment is the group of named objects which functions have access to. In other words, it is where the function will retrieve named objects used to execute the code from the body.

- The global environment is everything that we have loaded in, created ourselves or in base R.

- `ls()` tells you what you have created / assigned in the global environment (to get, set, test for and create environments, use `environment()`).

```{r}
#| label: global-environment

# ?environment()
ls()
y <- 1:5
ls()

```

- When calling functions, they have access to the names of objects in the namespace. It can map the name to where the object is stored. So it retrieves values through the names, so names are very powerful and why we can do computation work that is readable. 

```{r}
#| label: namespaces

# y is now in the namespace, so we have access to it
# -> the name gets mapped to its value
y*2

head(iris)

# -> we were able to find this because head and iris are a namespace in the global environment

```

- When we load a library, we are loading a namespace that has access to all the function names and dataset names in the library (package). Everything is a named object (objects, functions, etc.) and they are saved somewhere. The environment is where R finds the object.

Local namespaces

- When a function is called, a local namespace is created within the global environment (like a subspace of the global environment).

- When code is run in this local environment, R has access to all objects contained in the global environment.

- The primary difference is that when the local namespace is created, the formal argument names and their defined values are written into the local namespace. If these objects already exist in the global environment, they are overwritten locally while the function code is running. This is referred to as **name masking**.

- After the function code has completed running, the local environment is deleted. Thus, the local environment of a function is both nested in the global environment and is temporary.

```{r}
#| label: local-arguments

# when we call the mean function, x is defined locally but not globally
mean(x = y)

```

## Creating Functions

To create a function you must define the formals and body in using the `function()` function, and assign the definition to a name.

- Below we are giving a function as a value to the function name.

- Functions are first class objects (they store expressions, rather than data values).

- If there is a set of things we want to do, we can wrap a bunch of code into an expression by wrapping in curly brackets, then R thinks about it as one thing.

```{r}
#| label: defining-functions

# writing functions
function_name <- function(formal_arguments_go_here) {
  # code body expression goes here
}

formals(function_name)
body(function_name)
environment(function_name)

```

Also note, `function()` is a very weird function, since the body argument goes outside the parentheses! Technically even the example above is a real function (albeit a boring one).

Better example of simple function for adding up two numbers.

- Give functions a name that indicate what it does (intuitive names help a great deal with readable code).

- Style guide: Can use the assignment arrow `<-` for anytime something will be saved to the global environment (data values and functions), but within a function we can use `=` because it is only found locally and is temporary.

    - `<<-` Assigns values from a local environment (like a function call) and saves it in the global environment (this is a really bad idea because it could overwrite something that you already have in the global environment)
    
- Returning values: Functions implicitly return the last thing computed. So it is often a good idea to be explicit about what to return.

    - All standalone functions should have explicit returns. When doing an on-the-fly function in an `sapply()` for example, we can stylistically omit `return()` and still have readable code.

```{r}
#| label: defining-functions-assignment

# write function add two numbers
my_sum <- function(x, y){
  val = x + y
  return(val)
}
my_sum(1,2)

# bad idea
my_sum <- function(x, y){
  val <<- x + y
  return(val)
}
# val isn't in the global environment
my_sum(1, 2)

```

More complex functions: In a legitimate function, to have our function behave differently, we can use some if and else statements that are usually triggered by a yes / no or true / false.

`sapply()` is a function that calls other functions in an iterative way (really efficient for loop).

```{r}
# label: defining-functions-ifs

# now trying to do something cool (and / or breaking it)
# -> on rare occasions, add some random error to the sum
my_sum <- function(x, y, mischief = TRUE){
  if(mischief == FALSE){
    val = x + y
  } else {
    val = x + y + sample(c(-1,0,1), size = 1, prob = c(.02,.96,.02))
  }
  return(val)
}

# run the function 100 times using specified values for the arguments
# -> not using the X in our function call
sapply(X = 1:100, my_sum, x = 1, y = 2)

# equivalent to
for (i in 1:10) {
  print(my_sum(1, 2))
}

```

Can return something more complex than just a data value by returning lists. Functions that output lists are kinda the next step of complex computing ideas.

```{r}
# label: defining-functions-return-list

# modify function to return a list with multiple objects in it
my_sum <- function(x, y, mischief = TRUE){
  if(mischief == FALSE){
    val = x + y
  } else {
    val = x + y + sample(c(-1,0,1), size = 1, prob = c(.02,.96,.02))
  }
  return(list(value = val,
              mischief = mischief))
}

my_sum(1, 2)

```

### Calling a Function

To call a function, you simply use its name and define any required formal arguments. Typically this is done within a set of parentheses directly following the function name. Only a few special cases break from that pattern. 

What is more important to focus on is **if** and **how** we define the arguments. We may choose to define nothing in the arguments and the function might still work. 

```{r}
#| label: args-null
#| error: true

# write function that has no arguments to simply return this hardcoded value
no_arg <- function(){
  x = 10
  return(x)
}
no_arg()

# -> so it will generate an error if we tried to give it an argument that isn't defined in the function
no_arg(x = 20)

```

Typically these are made more flexible using dots `...`. All this does is put the values into the local namespace and continues running the function. These are optional arguments.

```{r}
#| label: args-dots

# define function wtih ... formal
no_arg2 <- function(...){
  x = 10
  return(x)
}

# functions works with no arguments
no_arg2() 

# and now with arguments
# -> it initializes the value of x, but then  overwrites the value of x internally
no_arg2(x = 20)

```

For "optional" arguments (arguments that sometimes matters and sometimes don't matter), we should define them as formal parameters with some default values. This way we can override them if we wanted to.

```{r}
#| label: args-defaults

# add default value to argument
# -> now it behaves typically how we want, but can change them if we want
no_arg3 <- function(x = 10){
  return(x)
}

# this is the behavior we typically will want
no_arg3()
no_arg3(x = 20)

```

The more important characteristics of function calls are related to how we define our arguments: implicitly and explicitly. If we rely on the default ordering of the arguements, then we are calling the arguments *implicitly*. If we refer to the argument by it's name while defining the values, then we are calling the argument *explicitly*. Often we use a mix of these methods, but it is important to be aware of how and why we choose to define our arguments.

A function runs through all of the named arguments, then goes in order for the unnamed arguments.

```{r}
#| label: arg_exp_implicit
#| error: true

# all of the following are equivalent
mean(x = 1:5, na.rm = TRUE)
mean(1:5, na.rm = TRUE)
mean(na.rm = TRUE, x = 1:5)

# the following is NOT equivalent
# -> it is expecting a numeric first
mean(TRUE, 1:5)

```

Possible convention: Implicitly give the function data values first, then use explicit calling for the options (this will still be readable because the function name will indicate what it's going to do with the data values).

### Automating procedures

The following example demonstrates code that would be better served by constructing a function to accomplish the task.

```{r}
#| label: example-change-to-functions-manual

# take the following olive oil data and standardize the columns
library(pdfCluster) # for the data
data(oliveoil) # ?oliveoil
head(oliveoil)

# start with a dataframe where create the z-scores for palmitic acid, then add the rest 
oo_standardized <- data.frame(
  palmitic = (oliveoil[,"palmitic"]-mean(oliveoil[,"palmitic"]))/sd(oliveoil[,"palmitic"])
)
head(oo_standardized)

# if wanted to do this manually for all columns we would do this
oo_standardized <- data.frame(
  palmitic = (oliveoil[,"palmitic"]-mean(oliveoil[,"palmitic"]))/sd(oliveoil[,"palmitic"]),
  palmitoleic = (oliveoil[,"palmitoleic"]-mean(oliveoil[,"palmitoleic"]))/sd(oliveoil[,"palmitoleic"]),
  stearic = (oliveoil[,"stearic"]-mean(oliveoil[,"stearic"]))/sd(oliveoil[,"stearic"]),
  oleic = (oliveoil[,"oleic"]-mean(oliveoil[,"oleic"]))/sd(oliveoil[,"oleic"]),
  linoleic = (oliveoil[,"linoleic"]-mean(oliveoil[,"linoleic"]))/sd(oliveoil[,"linoleic"]),
  linolenic = (oliveoil[,"linolenic"]-mean(oliveoil[,"linolenic"]))/sd(oliveoil[,"linolenic"]),
  arachidic = (oliveoil[,"arachidic"]-mean(oliveoil[,"arachidic"]))/sd(oliveoil[,"arachidic"]),
  eicosenoic = (oliveoil[,"eicosenoic"]-mean(oliveoil[,"eicosenoic"]))/sd(oliveoil[,"eicosenoic"])
)
head(oo_standardized)

```

Could do this (slighlty) better with a function, rather than copy and paste and change column names.

```{r}
#| label: example-change-to-functions-function

# write function to take a variable name and return the standardized column
oo_stand <- function(varname) {
  (oliveoil[,varname]-mean(oliveoil[,varname]))/sd(oliveoil[,varname])
}
oo_stand("palmitic")[1:10]

# building the standardized data almost as repetitive as previous approach
oo_standardized2 <- data.frame(
  palmitic =  oo_stand("palmitic"),
  palmitoleic =  oo_stand("palmitoleic"),
  stearic =  oo_stand("stearic"),
  oleic =  oo_stand("oleic"),
  linoleic =  oo_stand("linoleic"),
  linolenic =  oo_stand("linolenic"),
  arachidic =  oo_stand("arachidic"),
  eicosenoic =  oo_stand("eicosenoic")
)
head(oo_standardized2)

```

Next, can do this iteratively with for-loops to first print the result (for demonstration) and then to store the result.

```{r}
#| label: example-change-to-functions-for-loops

# intitialize names
acids <- names(oliveoil[3:10])

# loop over names and print first few standardized observations
for (acid_name in acids){
  print(oo_stand(acid_name)[1:5])
}

# initialization storage space for transformed values
oo_standardized3 <- matrix(data = rep(NA, nrow(oliveoil) * 8), 
                           nrow = nrow(oliveoil)) %>% 
  as.data.frame
names(oo_standardized3) <- names(oliveoil)[3:10]

# loop over the acid names
for (acid_name in acids){
  oo_standardized3[,acid_name] <- oo_stand(acid_name)
}
# check the outcome
head(oo_standardized3)

```

Best solution is to vectorize the procedure and use the APPLY type statements.

- `apply()` used to apply functions over the indices of an array.

- `lapply()` used to apply functions over the values in a list, outputs to a list.

- `sapply()` used to apply functions over the values in a list, outputs to a simplified array.

    - `lapply()` and `sapply(simplify = FALSE)` are the same thing, kinda how `paste()` and `paste0()` are the same except `paste0()` has a default separator (which is very useful, so they just gave it a shorthand function).

- `vapply()` used to apply functions over the values in a list, outputs to specified object type.

- `mapply()` used to apply functions over the corresponding values in multiple lists.

Karsten's Personal Use (with orders of magnitude >): `sapply` $>$ `lapply` $>>$ `mapply` $>>$ `vapply` $>>$ `apply`

```{r}
#| label: example-change-to-functions-apply

# loop over all column names with sapply()
# -> typically in practice the formals are given implicitly and just know that the object to be looped over is first and the function being applied is second
oo_better <- sapply(X = acids, FUN = oo_stand) %>% as.data.frame
head(better_oo)

# same thing, but return object is a list
oo_better_list <- lapply(X = acids, FUN = oo_stand)
str(oo_better_list)

```

Can use tidyverse equivalents from the `purrr` package ([documentation](http://purrr.tidyverse.org)), specifically `map()`.

- `map()` always returns a list, so this would be equivalent `lapply()`.

- Other variants include:

    - `map_vec()` must return a single value from each operation. If you know the output type, it is more efficient to use the corresponding variant such as `map_dbl()`, `map_chr()`, etc.
    
    - indexed map `imap()` applies a function to each element of a vector, and its index (which is shorthand for a particular use of `map2()` which iterates over 2 arguments at a time).
    
    - `lmap()` applies a function to list-elements of a list, useful for applying functions that take lists as arguments.
    
    - `pmap()` extends `map2()` to iterate over multiple arguments simultaneously.

```{r}
#| label: example-change-to-functions-map
#| error: true

# loop over all column names with sapply()
library(purrr)
#  
# help(package="purrr")
# ?map

# apply same function over names of columns using purrr equivalent function to lapply
oo_purrr_list <- map(.x = acids, .f = oo_stand)
str(oo_purrr_list)

# same thing, but now simplify the result
map_vec(.x = acids, .f = oo_stand)

# modify function to just return the st dev of the standardized columns
oo_stand_sd <- function(varname) {
  sd((oliveoil[,varname]-mean(oliveoil[,varname]))/sd(oliveoil[,varname]))
}
map_vec(.x = acids, .f = oo_stand_sd)

# know returning a double, so use map_dbl()
map_dbl(.x = acids, .f = oo_stand_sd)

```

Mini timing study to see which method of looping is most efficient.

- APPLY statements are quicker than for because they are run in C rather than R.

```{r}
#| label: timing

# create lots of lm from different random samples and store the slope of a particular variable
n <- 1000
betas <- rep(NA, n)

# initialize timing study and use for loop
timer_for <- Sys.time()
for(i in 1:n){
  betas[i] <- lm(palmitic ~ . , data = slice_sample(oliveoil, n = 100))$coefficients["oleic"]
}
Sys.time() - timer_for

# now use sapply()
# -> again not really using X = 1:n in the function that sapply() calls, it is just for the iterations to do the thing that many times
betas <- rep(NA, n)
timer_sapply <- Sys.time()
betas <- sapply(1:n, function(x) {
    lm(palmitic ~ . , data = slice_sample(oliveoil, n = 100))$coefficients["oleic"]
})
Sys.time() - timer_sapply

# now use map()
betas <- rep(NA, n)
timer_map <- Sys.time()
betas <- map(1:n, \(x) lm(palmitic ~ . , data = slice_sample(oliveoil, n = 100))$coefficients["oleic"])
Sys.time() - timer_map

# now use map_dbl()
betas <- rep(NA, n)
timer_dbl <- Sys.time()
betas <- map_dbl(1:n, \(x) lm(palmitic ~ . , data = slice_sample(oliveoil, n = 100))$coefficients["oleic"])
Sys.time() - timer_dbl

```

## Applying functions over multiple sets of parameters

### Simulate data 

In some cases we may wish to apply a complex function by putting in multiple sets of arguments and collecting the results. In the example below we explore a simple case where we want to gather simulated values from a beta distribution with several different shape parameter ($\alpha$, $\beta$) pairs.

```{r}
#| label: multiple-param-beta

# let's check what the rbeta() function does first in the help menu
# ?rbeta
# simulate n observations from a specific beta distribution
rbeta(5, shape1 = 1, shape2 = 1)

# want to apply the rbeta() function over all 100 unique combinations
# of shape1 = {0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5}
# and shape2 = {0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5}
param_combinations <- expand.grid(shape1 = seq(0.5, 5, by = 0.5),
                                  shape2 = seq(0.5, 5, by = 0.5))
head(param_combinations)

```

Could create two nested for loops to iterate over these pairs and store as it loops, but that is messy. Instead we will iterate over each pair using `mapply()` and `map2()`.

```{r}
#| label: iterate-multiple-args

# run simulation to generate 5 values beta values from each parameter combo
# use mapply() and simplify the result
set.seed(12345)
my_beta_sims <- mapply(FUN = rbeta, 
                       shape1 = param_combinations$shape1,
                       shape2 = param_combinations$shape2,
                       MoreArgs = list(n = 5),
                       SIMPLIFY = "array")

# map way shown at end

# rows in parameter storage correspond to the columns in the simulation array
param_combinations[1:10,]
my_beta_sims[,1:10]

# use map2()
# -> returns a list, so have to reduce it to turn into a matrix and then set the colnames to be readable
my_beta_sims <- param_combinations %>% 
  {map2(.$shape1, .$shape2,
       \(shape1, shape2) rbeta(n = 5, shape1 = shape1, shape2 = shape2))} %>% 
  reduce(cbind)
colnames(my_beta_sims) <- paste("sim", 1:nrow(param_combinations))

```

Or even better organization is to turn this into an 3-dimensional array where the index `[i,j,k]` represent the $i$th simulation, $j$th `shape1`, and $k$-th `shape2`.

```{r}
#| label: array

# convert copy of matrix to array (remember )
my_beta_sims_array <- my_beta_sims
dim(my_beta_sims_array) <- c(5,10,10)

# remember R goes down the columns first to fill the first layer
# -> then starts the next layer down the columns....
my_beta_sims_array[ , , 1] == my_beta_sims[, 1:10]

# display all simulations from the first shape 2 (columns = simulation and rows = iteration)
my_beta_sims_array[ , , 1]

```

### Summarize simulation 

Now suppose we want the mean for the simulation outcomes array. Use `apply()` to iterate over indices of the array.

```{r}
#| label: functions-on-array

# calc mean of a particular iteration
mean(my_beta_sims_array[ ,1 , 1])

# loop over each column in each layer (thus averaging over the rows) with apply()
beta_means <- apply(X = my_beta_sims_array,
                    MARGIN = c(2,3),
                    FUN = mean)
str(beta_means)

# map way requires different functions
# -> need to convert the array to a list first with array_tree() (a hierarchical list), and then map at the second level
# -> then simplify
# NOTE -> changing the order of dimensions in MARGIN transposes the result (i.e. order matters)
beta_means2 <- my_beta_sims_array %>% 
  array_tree(margin = c(3, 2)) %>% # first level list will be the layers, then the columns result$layer[[column]]
  map_depth(.depth = 2, .f = mean) %>% 
  reduce(cbind)
rownames(beta_means2) <- paste0("shape1_", unique(param_combinations$shape1))
colnames(beta_means2) <- paste0("shape2_", unique(param_combinations$shape2))

# compare two different methods
beta_means2 == beta_means

```

Now we can check our simulated means against a theoretical mean: $X \sim \text{beta}(\alpha, \beta) \Longrightarrow E(X) = \frac{\alpha}{\alpha + \beta}$.

```{r}
#| label: compute-means

# define function to calculate theoretical beta mean
theoretical_beta_mean <- function(alpha, beta){
  alpha / (alpha + beta)
}

# create a 10 x 10 matrix of the theoretical means for the beta distribution using the function above
theoretical_means <- mapply(FUN = theoretical_beta_mean,
                            alpha = param_combinations$shape1,
                            beta = param_combinations$shape2,
                            SIMPLIFY = "array")
dim(theoretical_means) <- c(10, 10)
theoretical_means

# now use map2, then simplify to vector and convert to matrix
theoretical_means2 <- param_combinations %>% 
  {map2(.$shape1, .$shape2, \(alpha, beta) theoretical_beta_mean(alpha, beta))} %>% 
  as_vector
dim(theoretical_means2) <- c(10, 10)

theoretical_means == theoretical_means2

# compare a few means
# -> note that for some reason beta_means2 is stored as a list but displaying like a matrix? but it is correct
theoretical_means[2,7] - beta_means2[2,7][[1]]
theoretical_means[1,10] - beta_means2[1,10][[1]]

```

### Entire workflow 

To confirm the simulated data is correct, we can do a bigger simulation (more iterations) and then compare to the theoretical means.

```{r}
#| label: entire-workflow

# steps of simulation
# -> simulate data based on parameter combinations
# -> keep in list form where each list all the iterations for a parameter pair
# -> then unpack and convert to 10 x 10 matrix
beta_simulation <- param_combinations %>% 
  {map2(.$shape1, .$shape2, \(shape1, shape2) rbeta(n = 1000, shape1 = shape1, shape2 = shape2))} %>% 
  map(mean) %>% 
  reduce(rbind)
dim(beta_simulation) <- c(10,10)

# compare simulated and theoretical means
theoretical_means - beta_simulation

```

## Writing Modular Functions

When you are writing functions to do a complex task, try to identify the system of simple tasks that are needed to accomplish it step-by-step. For example we might consider building functions to help us to automate the process of running stepwise selection based on AIC, then outputting a model object that we can use for prediction, or inference. 

We might initially think of one big function that can take a model matrix, X, and does a bunch of linear algebra to get beta estimates, calculates an AIC for the fitted model, then removes a columns from X to reduce the model... and repeat... a lot. Is that what we find inside the `step()` function? Run `body(step)` to see.

It isn't doing any of the linear algebra and AIC calculations we expected! The body of the step function is filled with, well more functions! It is function pinata!

What we find is that the `step()` function calls many other functions to do the individual smaller tasks that make up the complex procedure. If we search closely, there is a function for dropping a row, `drop1()`, within a `while` loop that keeps dropping rows until a condition is met. There is a function for updating the fit of the model, `update()`, and calculating the AIC, `extractAIC()`. There is even a function created at the start of the body that are used to help organize and return a helpful print statement about the stepwise procedure: `step.results()`. 

These functions are often times called "helper functions", completing the sub-tasks of the primary function. This modular format allows each function to complete a single simple task, and when used in combination, achieve a complex task. 

Thus, when modularizing code, the end goal to have one "overall" function that just has some helper functions within. On the way to having modularized code, is it better to start with an overall function and add modules sequentially (so e.g. if we call the overall, it will call module 1, so we check if that works; then add module 2 and call the overall which runs mod 1 and mod 2).... etc.

Below is an example (NOT RUN) to demostrate how helper functions can be used so that the overall function is body is relatively simple, all the work is being done by other functions. The overall function should be readable step by step.

```{r}
#| label: helper-functions
#| eval: false

# main data simulation function
# write a function to simulate a dataset for an n x p dataset for a regression model where q of the variables are inactive (signal variables)
make_sim_data <- function(n = 100, p = 20, q = 10, b = 0.1, sd_y = 1, sd_x = 1){
  
  # generate a n(iteration in sample) x p(sample) matrix from rnorm(0, sd_X)
  X = sapply(1:p, function(i) rnorm(n, 0, sd = sd_x))
  colnames(X) <- paste0("x",1:p)
  
  # specify population coefficients
  betas = c(rep(b, q), rep(0, p-q))
  
  # generate responses
  y = (X %*% betas)[,1] + rnorm(n,0, sd_y)
  
  # combine data
  sim_data <- data.frame(y,X)
  
  return(sim_data)

}

# helper functions for running the experiment with variable selection

# function for choosing with stepwise and fitting a regression
step_var_mod <- function(df){
  step_selected <- step(lm(y ~ . , data = df), trace = FALSE)
  return(step_selected)
}

# function for choosing with lasso and fitting regression
lasso_var_mod <- function(df){
  
  cv.out = cv.glmnet(x = as.matrix(x = df[ ,-which(names(df) == "y")]),
                      y = df$y, alpha = 1, type.measure = "deviance")
  
  lasso_mod = glmnet(x = as.matrix(x = df[, -which(names(df) == "y")]),
                      y = df$y, alpha=1, lambda = cv.out$lambda.1se)
  
  lasso_vars = names(lasso_mod$beta[,1])[which(lasso_mod$beta[,1] != 0)]
  
  if (length(lasso_vars) == 0) lasso_vars <- names(lasso_mod$beta[,1])[1]
  
  lasso_selected = lm(formula(paste0("y ~ 1 + ", paste(lasso_vars, collapse = " + "))), data = df)
  
  return(lasso_selected)
  
}

# function for finding number of variables included
select_var_count <- function(lin_mod){
  length(coef(lin_mod))-1
}

# function for finding 10-fold cross validated RMSE
select_cv_rmse <- function(lin_mod){
  
  cv_result = train(formula(lin_mod), 
    data = lin_mod$model,
    method = "lm",
    trControl = caret::trainControl(method = "cv", number = 10),
    tuneGrid = data.frame(intercept=TRUE))
  
  return(cv_result$results$RMSE)
  
}

# function to run trial 
run_trial <- function(selection_alg, df){
  
  start_time = Sys.time()
  
  tmp_mod = selection_alg(df) # not sure where this function comes from
  
  end_time = Sys.time()
  
  return(data.frame(var_count = select_var_count(tmp_mod),
                    rmse = select_cv_rmse(tmp_mod), # nor this function
                    time = difftime(end_time, start_time, units = "secs")))
  
}

# make into a function of n_sims, n, p and q
sim_var_select <- function(n_sim = 10, n = 100, p = 10, q = 5, var_select_ftn = step_var_mod){
  
  results = NULL
  
  for(i in 1:n_sim){
    
    sim_data <- make_sim_data(n = n, p = p,q = q)
    
    results <- rbind(results, run_trial(var_select_ftn, sim_data))
    
  }
  
  return(results)
}

```

